Estimation of the perceived time of presence of sources in urban acoustic environments using deep learning techniques

### 2019-09-19 paper/paperV2.tex
FG proof v2

The impact of urban sound on human beings has often been extensively studied from a negative point of view (sound pollution). In the two last decades, the interest of studying its positive impact has been revealed with the soundscape approach (resourcing spaces). The literature shows that the recognition of sources plays a great role in the way humans are affected by sound environments. There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as their perceived time of presence.

@@

The considered indicators are compared to subjective assessments obtained during the listening test to identify their capacity to explain dimensions of soundscape perception. The analysis is performed using the arithmetic mean of subjective assessments over all participants for replicated and simulated scenes. The source-specific sound level $L_{eq, s}$, emergence $\Delta L_{s}$ and estimated time of presence $\hat T_s(\alpha, \beta)$ indicators cannot be directly computed on the 6 recorded scenes, as their ground truth source composition is unknown. Thus, these 6 scenes are excluded of the study. Two simulated scenes containing only one source are also excluded because of implementation concerns, resulting in $n=92$ scenes.  TODO implementation : more details needed

@@

In the context of urban soundscape quality assessment, pleasantness emerges as the first affective dimension~\cite{axelsson2010}. Pleasantness has been modeled from both the perceived overall loudness in a scene~\cite{blauert1997, jekosch2004} and its content in terms of active sources~\cite{axelsson2010, lavandier2006, ricciardi2014, aumond2017}. To demonstrate the potential of the proposed method for predicting the time of presence of sources is applied to pleasantness estimation.

@@

Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{1, \varphi}$ model presented in Section~\ref{sec:app_mdls}. Thus, only the outputs corresponding to the presence of birds are used. TODO The results are compared to estimates using ground truth $\hat T_s(\alpha_{opt}, \beta_{opt})$ labels in Table~\ref{tab:pppred} TODO reformuler. First, pleasantness estimations are equally effective using the model's predictions or the ground truth labels, with about 0.84 RMSE and 82\% of explained variance on the perceptual experiment corpus. This performance of the detection model is expected given the low errors on time of presence estimates in Table~\ref{tab:perf_cmp}. Labels from the detection model result in lower errors on the first sub-corpus containing replicated scenes with sources not seen during training. This result may indicate that the detection model generalizes well to additional sources, though a larger sample size would be required to confirm this interpretation. For the simulated subcorpus of the experiments all performance metrics are comparable. Applying the detection model on the corpus of recorded scenes for which ground truth pleasantness is available results in a RMSE of 1.09 and a decrease in $R^2_{adj}$. This is expected as these scenes are the most distant from the training corpus in terms of sources and scenarios.

@@

The methodology proposed in this paper can be extended to include additional sound sources provided that enough excerpts of isolated source occurrences are available. Indeed, the scene simulation process, the time of presence indicator, and the deep learning model are all independent from the considered taxonomy. Of course, the complexity of the learning process scales with the number of considered simultaneously active sources. Particularly, differentiating between sources with similar spectral shapes may yield higher error rates as the deep learning architecture relies on the identification of patterns in third octave spectra. To obtain sufficient amounts of data, data augmentation techniques can be used to diversify the training dataset by applying slight modifications to existing examples, such as filtering, pitch shifting or time stretching. TODO cite may be paper of dcase on data augmentation ?

>>>>

The impact of urban sound on human beings has often been studied from a negative point of view (sound pollution). In the two last decades, the interest of studying its positive impact has been revealed with the soundscape approach (resourcing spaces). The literature shows that the recognition of sources plays a great role in the way humans are affected by sound environments. There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as their perceived time of presence.

@@

The considered indicators are compared to subjective assessments obtained during the listening test to identify their capacity to explain dimensions of soundscape perception. The analysis is performed using the arithmetic mean of subjective assessments over all participants for replicated and simulated scenes. The source-specific sound level $L_{eq, s}$, emergence $\Delta L_{s}$ and estimated time of presence $\hat T_s(\alpha, \beta)$ indicators cannot be directly computed on the 6 recorded scenes, as their ground truth source composition is unknown. Thus, these 6 scenes are excluded of the study. Two simulated scenes contained only one source, leading to infinite source emergences. Statistics including these scenes cannot be computed and they are thus excluded in this analysis, which as a result includes $n=92$ scenes.

@@

In the context of urban soundscape quality assessment, pleasantness emerges as the first affective dimension~\cite{axelsson2010}. Pleasantness has been modeled from both the perceived overall loudness in a scene~\cite{blauert1997, jekosch2004} and its content in terms of active sources~\cite{axelsson2010, lavandier2006, ricciardi2014, aumond2017}. To demonstrate the potential of the proposed method for predicting the time of presence of sources, it is applied to pleasantness estimation.

@@

Pleasantness predictions are obtained by substituting time of presence estimates computed from the presence detection architecture's outputs to the $\hat P_{1, \varphi}$ model presented in Section~\ref{sec:app_mdls}. Thus, only the outputs corresponding to the presence of birds are used. Table~\ref{tab:pppred} presents the performance of pleasantness estimations using outputs from the deep learning architecture compared to those using ground truth $\hat T_s(\alpha_{opt}, \beta_{opt})$ labels computed with eq.~\ref{eq:ts} on separated source-specific channels. First, pleasantness estimations are equally effective using the model's predictions or the ground truth labels, with about 0.84 RMSE and 82\% of explained variance on the perceptual experiment corpus. This performance of the detection model is expected given the low errors on time of presence estimates in Table~\ref{tab:perf_cmp}. Labels from the detection model result in lower errors on the first sub-corpus containing replicated scenes with sources not seen during training. This result may indicate that the detection model generalizes well to additional sources, though a larger sample size would be required to confirm this interpretation. For the simulated subcorpus of the experiments all performance metrics are comparable. Applying the detection model on the corpus of recorded scenes for which ground truth pleasantness is available results in a RMSE of 1.09 and a decrease in $R^2_{adj}$. This is expected as these scenes are the most distant from the training corpus in terms of sources and scenarios.

@@

The methodology proposed in this paper can be extended to include additional sound sources provided that enough excerpts of isolated source occurrences are available. Indeed, the scene simulation process, the time of presence indicator, and the deep learning model are all independent from the considered taxonomy. Of course, the complexity of the learning process scales with the number of considered simultaneously active sources. Particularly, differentiating between sources with similar spectral shapes may yield higher error rates as the deep learning architecture relies on the identification of patterns in third octave spectra. To obtain sufficient amounts of data, data augmentation techniques can be used to diversify the training dataset by applying slight modifications to existing examples, such as filtering, pitch shifting or time stretching~\cite{salamon2017-2, lasseck2018}.
